\chapter*{Acknowledgment}

I would like to thank Erik Meijer, my supervisor during this thesis project, for all his support, enthusiasm and hacker mentality. \textit{Erik, it was an honor to receive an email in May 2014, asking whether I was interested in doing my master thesis with you. From the beginning it was clear to both of us that Reactive Programming and Rx were a shared interest. It was great working with you on RxMobile, which has turned out to be very useful for this thesis. You really inspired me to have a hacker mentality, always keep trying to get things working and to never give up. I sincerely hope the completion of this thesis will not be the \code{onCompleted} event on the stream of our collaborations, but instead will just be the \code{onNext} event of a finished project on a much longer stream!}

I also want to thank Georgios Gousios and Arie van Deursen for taking their time to be part of my thesis committee and providing me with great feedback.

Furthermore, I would like to thank my fellow students and friends at TU Delft, Eddy Bertoluzzo, Georgi Khomeriki, Mircea Voda, Mike de Waard and Lars Willems, who were doing their master theses in parallel with me. Thanks for the great time we had together at EWI-HB 08.250 and the many discussions at the whiteboard and at Hangouts/Skype!

A special thanks to Michel van Heest, my brother, for creating all the diagrams in this thesis.

Finally, I would like to thank my family, friends and colleagues for their support and interest in my thesis work, even though most of them still don't understand a word of what will follow in the rest of this thesis!

\begin{flushright}
Richard van \textsc{Heest}\\
\textit{Middelharnis, The Netherlands}\\
\textit{\monthyeardate\today}
\end{flushright}