\chapter{Introduction to feedback control}
Control is something we deal with on a daily basis. Whether it is the temperature in our houses, the number of people in line at the supermarket's checkout or something more critical like the number of neutrons emitted during a nuclear reaction in a reactor, we have to control it to avoid potential chaos.

In software development we too want to have control over the products we both create and use. For example, most applications have a settings menu where the user can alter the application's behavior to their liking; to ensure our code to be working correctly we use compilers, IDEs and automated testing tools that help us trace errors and bugs; companies like Google, Facebook and Twitter do A/B-testing and other kinds of user studies when introducing new features and with that control and improve the usability of their product; and last but not least, in cloud computing we want to control the number jobs in a queue and scale up or down depending on the amount of jobs in the queue, such that all jobs get executed as fast as possible with spending the least amount of resources.

A common factor in controlling a system is the notion of feedback: you change something in the environment and see how well it adjusts towards your desired outcome. If the lines at checkout get too long, most likely an extra cash register needs to be opened; if your users don't perform well on an A/B-test of your latest features, you should probably revise these features rather than bringing them into production; and when you alter the settings in an application, you check whether the new behavior is more to your desire and maybe fiddle around with them some more until it is more to your liking.

Controlling a system without the notion of feedback is possible, however this requires an exact model of the system under control. External forces that are not part of the model cannot be allowed to disturb the system. In most cases this model does unfortunately not exist and control has to incorporate the system's output to come up with it's next input. If the output is not taken into account, the system can drift off and end up in undesired situations. A nice example of this are Microsoft's user studies: the user's feedback was not taken into account when Office 2007 introduced the ``ribbon'' to replace the existing user interface nor when the start button disappeared in Windows 8 and suddenly every Windows user lost their 20+ years of accumulated muscle memory for commanding the Windows operating system \cite{meijer2014-embracing-the-hacker-way}.

In physics and engineering, feedback control is a commonly used technique and is also fully backed by mathematics. A series of equations and theorems such as Laplace Transforms and differential equations describe the abstract behavior of a feedback controlled system and allow to evaluate its outcome after running it for a certain time and with a certain input \cite{hellerstein2004-feedback, janert2013-feedback}. These theorems and equations are usable in science because everything else is described in the `language of mathematics' as well. We have known the differential equations to Newton's law of cooling, damped harmonic oscillators and so forth for centuries, as well as their Laplace Transforms, which are needed for the `feedback equations'.

This is all in contrast to the current situation in computer science. Although proven to be very effective in physics, computer science has not yet widely adopted the techniques of feedback control. Instead, complex algorithms are written to control a web cache or do cloud scaling. Although it seems very naturally applicable, it is surprising to find that such a simple and effective technique is mostly ignored.

A potential reason for this lack of using feedback control lies in the fact that computer science does not yet fully understand the whole mathematical background of the datastructures and applications that are being used and created. Given this lack of understanding, we are not yet able to describe to which laws our datastructures and application obey. What would be the equivalent of Newton's laws for a web cache \cite{janert2013-feedback}? Only recently some work has been done on understanding the mathematics behind this \cite{beckmann2015-cache-calculus} and we may very well see more of this in the future, which will eventually enable computer science to create a more formal mathematical model with theorems and equations equivalent to physics.

The lack of applying feedback control in computer science is also reflected in this technique commonly not being taught in university degree computer science education. Although it is one of the standard courses in physics and engineering studies, it is not at all present in the computer science curricula. And in the rare occasions it is taught, it is done in a way that a physics student would learn about it, namely as yet another maths course, even though this way of treating feedback control in computer science is not really applicable!

In this chapter we will introduce the basic concepts of feedback control with as little mathematics as possible. We will start with an overview of what a feedback system consists of and how it works. Next we will go into some detail on how to control a feedback system. Finally we will introduce a simple toy example that demonstrates the power of feedback control and translates these concepts into an imperative style program.

\input{3.1-overview}

\input{3.2-pid-control}

\input{3.3-extended-example}
